DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-26 Paper_May_2012/main.tex
Kommentare hinzugefuegt, mein Beispiel weiter aufgeraeumt, Einrueckung im Code an das neue, schmalere Design angepasst. Es gibt noch Probleme mit texttt im Text, TeX trennt diese Woerter nicht automatisch. Wollen wir Code-fragmente ueberhaupt trennen lassen?

\institute{F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}   
}

@@

In order to solve this problem, i.e., calculate a sequence of values $J(u_h)$ approximating $J(u)$,
we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename TRIANGULATION & triag, 
           typename FE & fe>
      void SpaceTimeHandler(Triangulation &, FE);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

Then, the two characteristic features \texttt{LocalPDE} and \texttt{LocalFunctional} objects come into play:
\begin{lstlisting}

  LocalPDE local_pde;
  LocalFunctional local_functional;

\end{lstlisting}
These hold the information of the PDE and the functional under consideration. Further explication is given below because
the definition of this classes is done separately in \texttt{localpde.h} resp.   \texttt{localfunctional.h}.

In the next step, we define an object of type \texttt{SpaceTimeHandler}, which handles all the things connected with the triangulation as well as the finite element space. This class is also capable of organizing degrees of freedom in time-dependent problems, hence the name. 
\begin{lstlisting}

  SpaceTimeHandler dof_handler(triangulation, finite_element);

\end{lstlisting}
There exist different versions of this class for forward as well as optimization problems.

Now we have everything at hand to create an instantiation of type \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{local\_pde} and the \texttt{dof\_handler}.
\begin{lstlisting}

  PDEProblemContainer<LocalPDE> prob_container(local_pde,
                                               dof_handler);

\end{lstlisting}
After that, we add the functional under consideration and, because it is a boundary functional, we specify the part of the boundary on which the functional operates (this is done via a 'boundary color', $1$ represents $\Gamma_{circ}$).
\begin{lstlisting}

  prob_container.AddFunctional(&local_functional);
  prob_container.SetBoundaryFunctionalColors(1);

\end{lstlisting}
Next, we incorporate the Dirichlet data, which is given by the class
\texttt{DirichleData}. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color in a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}

  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0, comp_mask, dd);

\end{lstlisting}
Now we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}

@@

\end{lstlisting}
Here, \texttt{NEWTON} itself is an abbreviation of
\begin{lstlisting}

  typedef NewtonSolver<LINSOLVE> NEWTON;

\end{lstlisting}
The  template parameter \texttt{LINSOLVE} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}

  typedef DirectLinearSolver<MATRIX, VECTOR> LINSOLVE;

\end{lstlisting}
With the type settled, we create the object \texttt{solver}, which needs the discretization (represented here by the \texttt{prob\_container}), the parameter reader as well as the quadrature rules.
\begin{lstlisting}

  PDE pde(prob_container, pr,
                    quad_formula, face_quad_formula);

\end{lstlisting}
Finally, we compute the functional $J$ by calling:
\begin{lstlisting}

  pde.ComputeReducedFunctionals();

\end{lstlisting}

@@

\begin{lstlisting}
  void CellEquation(const CellDataContainer &cdc,
                    VECTOR &local_cell_vector)
    {
      const auto& fe_values = cdc.GetFEValuesState();
      unsigned int n_dofs_per_cell = cdc.GetNDoFsPerCell();
      unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
  \begin{lstlisting}
     cdc.GetValuesState("last_newton_solution", uvalues);
     cdc.GetGradsState("last_newton_solution", ugrads);
 \end{lstlisting}
 In these two lines, we load the (gradient of the) solution of the last newton iteration in all the quadrature points on the cell into the two vectors \texttt{uvalues} and \texttt{ugrads}.

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{h}^i$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<n_dofs_per_cell;i++)
        local_cell_vector(i) += 
           ( _nu * vgrads * phi_i_grads_v 
             + //all the other terms)
             * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{CG and WW: Pruefen, ob der folgende Absatz nun gut ist:}
The main difference is that as we only want to evaluate a functional and not a
residual-vector, we just loop over the quadrature points and not over the
degrees of freedom on the face. Another deviation, special to the case at
hand, is that we do not have to specify the method \texttt{CellEquation}, but
\texttt{BoundaryValue} instead because $J$ is just number to be computed. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$).

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l(\phi)$.
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES & times, 
           typename TRIANGULATION & triag, 
           typename FE & fe>
   void SpaceTimeHandler(Times &, Triangulation &, FE &);
\end{lstlisting}

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
\todo{CG and WW: bitte pruefen}
After choosing the time-stepping scheme, the SpaceTimeHandler  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality to solve stationary PDEs. 
Now it is apparent, why we introduced the \texttt{StatPDEProblem}, because
we define the \texttt{InstatPDEProblem} to contain the necessary 
loop over all time points. Consequently, the implementation 
of nonstationary problems is quite similar to stationary problems,
which was one major goal we claimed at the beginning.
Next, we will describe the methods (i.e., changes) the user needs to 
provide in addition to the stationary case.

@@

It is well-known in fluid mechanics that the pressure and incompressibility terms should be 
treated fully implicitly when time-discretizing 
the Navier-Stokes equations.
For this reason, the implementation is split up
into several items to account for fully implicit 
terms and remaining terms for which we need 
old time-step and present time-step values. 
Hence, in the local cell equation, the parameters
\texttt{scale} and \texttt{scale\_ico}
are used to distinguish these terms:
\begin{lstlisting}
  void CellEquation(...)
   {
   // Usual loops over DoFs and quadrature points
     {
      local_cell_vector(i) += 
          scale * 
          // terms like (\nabla v^{n+1}, \nabla \phi) etc.
        + scale_ico *
          // terms like (-p^{n+1},\nabla\cdot \phi)
     }
   }
\end{lstlisting}
Analogous procedure is performed in the local 
cell matrix.
\begin{remark}
Note that in the stationary case \dope{} automatically sets $ \texttt{scale\_ico} = \texttt{scale}$, so that in fact the user has to implement the \texttt{CellEquation} only once and can henceforth reuse it in all his stationary an instationary problems. This transfers of course also to other methods like \texttt{CellMatrix}.
\end{remark}

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in figure~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
\mathcal {J}( u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\todo{For CG: Bildunterschrift und die $2$ am Ende und 
das cal J}
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.8\textwidth]{Pictures/local_grid_NS.png}}
\caption{Caption.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $\mathcal J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $\mathcal J$. The price we have to pay ist the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In figure~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\ref{fig:NS_local_grid}
\begin{figure}[hbt]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/ConvergenceNS}}
\caption{Comparison of global and local mesh refinement with respect to the error in the drag-functional $\mathcal J$.}
\label{fig:NS_comparison}
\end{figure}

@@

\section{Conclusions}
\todo{Frage an CG und Winni: Guter Abschluss?}
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems. The key idea behind 
the algorithms is the computation of some target 
functional. This is achieved by unified interface 
to state-of the art time-stepping methods, nonlinear solvers and optimization
routines. The 
capabilites of DOpElib are demonstrated
wit the help of three examples.

>>>>

\institute{Christian Goll \at
              Institute of Applied Mathematics, University of Heidelberg
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              Tel.: +49 622-154-6154\\
              \email{cgoll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}   
}

@@

In order to solve this problem, i.e., calculate a sequence of values $J(u_h)$ approximating $J(u)$,
we need to discretize 
the equation. In \dope, we are using finite elements to do this. This means we need to pick a 
subdivision $\mathcal T_h$, for instance into quadrilaterals,
of the domain $\Omega$ and a finite element. In the concrete case, we could, 
for instance, take the $\mathcal Q_2/\mathcal Q_1$ Taylor-Hood element.
As it is clear, that the initialization of the unknowns, the local shape 
functions, removal of hanging node constraints, etc. always follows the 
same procedure, we define an object which takes the finite element (FE)
and the triangulation and takes care of all other things related to
handling the finite element space $V_h$. We will refer, to 
this as the \texttt{SpaceTimeHandler}. Then, it is clear that the simplest 
form of the constructor should have the following interface:
\begin{lstlisting}
  template<typename TRIANGULATION, typename FE>
     void SpaceTimeHandler(TRIANGULATION& triag, FE& fe);
\end{lstlisting}
Next, one needs to solve the discrete equation
\begin{align}\label{eq:discrete_equation}
a(u_h)(\phi_h) = (f,\phi_h) \quad \forall \phi \in V_h.
\end{align}
To this end, one can employ a newton type method with some globalization 
like line-search. To do so, one needs to calculate the residual of the 
equation, as well as to solve linearized equations for the calculation of 
the newton method and then employing some kind of globalization strategy.
This again will usually require the evaluation of the residual.

@@

Then, the two characteristic features \texttt{LocalPDE} and \texttt{LocalFunctional} objects come into play:
\begin{lstlisting}
  LocalPDE local_pde;
  LocalFunctional local_functional;
\end{lstlisting}
These hold the information of the PDE and the functional under consideration. Further explication is given below because
the definition of this classes is done separately in \texttt{localpde.h} resp.   \texttt{localfunctional.h}.

In the next step, we define an object of type \texttt{SpaceTimeHandler}, which handles all the things connected with the triangulation as well as the finite element space. This class is also capable of organizing degrees of freedom in time-dependent problems, hence the name. 
\begin{lstlisting}
  SpaceTimeHandler dof_handler(triangulation,
                               finite_element);
\end{lstlisting}
There exist different versions of this class for forward as well as optimization problems.

Now we have everything at hand to create an instantiation of type \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{local\_pde} and the \texttt{dof\_handler}.
\begin{lstlisting}
  PDEProblemContainer<LocalPDE> 
                      prob_container(local_pde,
                                     dof_handler);                                            
\end{lstlisting}
After that, we add the functional under consideration and, because it is a boundary functional, we specify the part of the boundary on which the functional operates (this is done via a 'boundary color', $1$ represents $\Gamma_{circ}$).
\begin{lstlisting}
  prob_container.AddFunctional(&local_functional);
  prob_container.SetBoundaryFunctionalColors(1);
\end{lstlisting}
Next, we incorporate the Dirichlet data, which is given by the class
\texttt{DirichleData}. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color in a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}
  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0,
                                            comp_mask,
                                            dd);
\end{lstlisting}
Now we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}

@@

\end{lstlisting}
Here, \texttt{NEWTON} itself is an abbreviation of
\begin{lstlisting}
  typedef NewtonSolver<LINSOLVE> NEWTON;
\end{lstlisting}
The  template parameter \texttt{LINSOLVE} describes which linear solver we use in this example. We opted for a direct solver, so: 
\begin{lstlisting}
  typedef DirectLinearSolver<MATRIX, VECTOR> LINSOLVE;
\end{lstlisting}
With the type settled, we create the object \texttt{solver}, which needs the discretization (represented here by the \texttt{prob\_container}), the parameter reader as well as the quadrature rules.
\begin{lstlisting}
  PDE pde(prob_container, pr,
                    quad_formula, face_quad_formula);
\end{lstlisting}
Finally, we compute the functional $J$ by calling:
\begin{lstlisting}
  pde.ComputeReducedFunctionals();
\end{lstlisting}

@@

\begin{lstlisting}
  void CellEquation(const CellDataContainer &cdc,
                    VECTOR &local_cell_vector)
  {
    const auto& fe_values = cdc.GetFEValuesState();
    unsigned int dofs_per_cell = cdc.GetNDoFsPerCell();
    unsigned int n_q_points = cdc.GetNQPoints();
\end{lstlisting}
 The \texttt{CellDataContainer} contains all the information we need to compute the cell-contributions (things like number of quadrature points, number of degrees of freedom on the cell, the \texttt{dealii::FEValues}-objects, etc.).
  \begin{lstlisting}
    cdc.GetValuesState("last_newton_solution", uvalues);
    cdc.GetGradsState("last_newton_solution", ugrads);
 \end{lstlisting}
 In these two lines, we load the (gradient of the) solution of the last newton iteration in all the quadrature points on the cell into the two vectors \texttt{uvalues} and \texttt{ugrads}.

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{h}^i$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<dofs_per_cell;i++)
        local_cell_vector(i) += 
           ( _nu * vgrads * phi_i_grads_v 
             + //all the other terms)
             * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{CG and WW: Pruefen, ob der folgende Absatz nun gut ist:}
The main difference is that as we only want to evaluate a functional and not a
residual-vector, we just loop over the quadrature points and not over the
degrees of freedom on the face. Another deviation, special to the case at
hand, is that we do not have to specify the method \texttt{CellEquation}, but
\texttt{BoundaryValue} instead because $J$ is just a number to be computed. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face-wise (for the faces that lie on $\Gamma_{circ}$).

@@

\subsection{Solving Nonstationary PDEs}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form $a(u)(\phi)$ and some
right hand side functional $l(\phi)$.
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi)\, dt = \int_I l(\phi)\, dt \quad \forall\Phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we extend the previous example and consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\begin{align*}
&\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt\\
&\quad + (v(0) - v^0, \phi(0))
= \int_I l(\phi) \, dt 
\end{align*}
for all $\Phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename TIMES, typename TRIANGULATION, 
           typename FE>
   void SpaceTimeHandler(TIMES& times,
                         TRIANGULATION& triag,
                         FE& fe);
\end{lstlisting}

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
\todo{CG and WW: bitte pruefen}
After choosing the time-stepping scheme, the SpaceTimeHandler  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality \todocg{we employed} to solve stationary PDEs.\todocg{Hier Beispiele, was genau wir wiederverwenden? Oder auf den naechsten Abschnitt verweisen?} 
Now it is apparent, why we introduced the \texttt{StatPDEProblem}, because
we define the \texttt{InstatPDEProblem} to contain the necessary 
loop over all time points. Consequently, the implementation 
of nonstationary problems is quite similar to stationary problems,
which was one major goal we claimed at the beginning.
Next, we will describe the methods (i.e., changes) the user needs to 
provide in addition to the stationary case.

@@

It is well-known in fluid mechanics that the pressure and incompressibility terms should be 
treated fully implicitly when time-discretizing 
the Navier-Stokes equations.
For this reason, the implementation is split up
into several items to account for fully implicit 
terms and remaining terms for which we need 
old time-step and present time-step values. 
Hence, in the local cell equation, the parameters
\texttt{scale} and \texttt{scale\_ico}
are used to distinguish these terms:
\begin{lstlisting}
  void CellEquation(...)
   {
   // Usual loops over DoFs and quadrature points
     {
      local_cell_vector(i) += 
         scale * 
         // terms like (\nabla v^{n+1}, \nabla\phi) etc.
        + scale_ico *
         // terms like (-p^{n+1},\nabla\cdot\phi)
     }
   }
\end{lstlisting}
Analogous procedure is performed in the local 
cell matrix.
\begin{remark}
Note that in the stationary case \dope{} automatically sets $ \texttt{scale\_ico} = \texttt{scale}$, so that in fact the user has to implement the \texttt{CellEquation} only once and can henceforth reuse it in all his stationary an instationary problems. This transfers of course also to other methods like \texttt{CellMatrix}.
\end{remark}

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in figure~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay ist the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In figure~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\begin{figure}[hbt]
\centering
\resizebox{0.5\textwidth}{!}{\input{Pictures/ConvergenceNS}}
\caption{Comparison of global and local mesh refinement with respect to the error in the drag-functional $J$.}
\label{fig:NS_comparison}
\end{figure}

@@

\section{Conclusions}
\todo{Frage an CG und Winni: Guter Abschluss?}\todocg
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems.  The key idea behind 
the algorithms is the computation of some target 
functional \todocg{Ich wuerde eventuell die userfreundlichkeit mehr in den Fokus ruecken? The key idea of the software is to supply a toolkit which allows for fast implementation of the problems at hand, but at the same time gives the user full flexibility to change or alter the used algorithms}. This is achieved by a unified interface 
to state of the art time-stepping methods, nonlinear solvers and optimization
routines. The capabilites of DOpElib are demonstrated with the help of three examples.
